The finite representation property for some reducts of relation algebras

### 2021-01-23 frp-paper/Paper.tex
wip

The logic of such structures is the Lambek calculus \cite{lambek1958mathematics} allowing one to characterise inference in categorial grammars, the equivalent version of context-free grammars \cite{pentus1993lambek}. One may define the Lambek calculus as the following (cut-free) sequent calculus:

\begin{definition}
  $ $

  \begin{prooftree}
  \AxiomC{$ $}
  \RightLabel{\scriptsize{ax}}
  \UnaryInfC{$\varphi \vdash \varphi$}
  \end{prooftree}

\begin{minipage}{0.5\textwidth}
  \begin{flushleft}
        \begin{prooftree}
      \AxiomC{$\Gamma \vdash \varphi$}
      \AxiomC{$\Delta, \psi, \Theta \vdash \theta$}
      \RightLabel{$\backslash \vdash $}
      \BinaryInfC{$\Delta, \Gamma, \varphi \backslash \psi, \Theta \vdash \theta$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \varphi$}
      \AxiomC{$\Delta, \psi, \Theta \vdash \theta$}
      \RightLabel{$/ \vdash $}
      \BinaryInfC{$\Delta, \psi \: / \: \varphi, \Gamma, \Theta \vdash \theta$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Gamma, \varphi, \psi, \Delta \vdash \theta$}
      \RightLabel{$\bullet \vdash $}
      \UnaryInfC{$\Gamma, \varphi \bullet \psi, \Delta \vdash \theta$}
    \end{prooftree}
\end{flushleft}
\end{minipage}
\begin{minipage}{0.5\textwidth}
  \begin{flushright}
       \begin{prooftree}
      \AxiomC{$\varphi, \Pi \vdash \psi$}
      \RightLabel{$\vdash \backslash$}
      \UnaryInfC{$\Pi \vdash \varphi \backslash \psi$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Pi, \varphi \vdash \psi$}
      \RightLabel{$\vdash /$}
      \UnaryInfC{$\Pi \vdash \varphi \: / \: \psi$}
    \end{prooftree}

    \begin{prooftree}
      \AxiomC{$\Gamma \vdash \varphi$}
      \AxiomC{$\Delta \vdash \psi$}
      \RightLabel{$\vdash \bullet$}
      \BinaryInfC{$\Gamma, \Delta \vdash \varphi \bullet \psi$}
    \end{prooftree}
\end{flushright}
\end{minipage}
\end{definition}

@@

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $|\mathcal{A}| = \omega$ and $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
  \end{enumerate}
\end{lemma}

@@

  \begin{enumerate}
    \item Let $h$ be a representation of $\mathcal{A}$ over some domain $D$. $\forall$ plays $(a, b)$ with $a \not\leq b$ for the initial round. Let $N = \langle E, V, l \rangle$ be a network being played.

    TODO: define an embedding $\iota : E \to D$ and check the presence of a winning strategy explicitly

    \item Let $a, b \in \mathcal{A}$ such that $a \not\leq b$. We consider a play of the game where $\forall$ pick the pair $(a, b)$ in the initial round and plays $(\mathcal{N},x, y, z, c, d)$ in the further rounds for each $x, y, z \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$. Here, $c \in l_N(x, y)$ and $d \in l_N(y, z)$.

@@

\item Suppose $(x, y) \in rep(a) ; rep(b)$. We need $(x, y) \in rep(a;b)$. That is, there exists $z \in \mathcal{Nodes}(\mathcal{N})$ with $(x, z) \in rep(a)$ and $(z, y) \in rep(b)$. So, there are $c, d$ such that $c \leq a$ with $c \in l_{\mathcal{N}}(x, z)$ and $d \leq b$ with $d \in l_{\mathcal{N}}(z, y)$.
We also know that $l_{\mathcal{N}}(x, z) ; l_{\mathcal{N}}(z, y) \subseteq l_{\mathcal{N}}(x, y)$ because all composition moves have been played.
So $c ; d \in l_{\mathcal{N}}(x, y)$. That makes $(x, y) \in rep(a;b)$ since $c ; d \leq a ; b$.

\item Injectivity: TODO.
\end{itemize}
  \end{enumerate}
\end{proof}

@@

\begin{lemma}
  For each $n < \omega$ there exists a first-order sentence $\sigma_n$ such that $\exists$ has a winning strategy in $\mathcal{G}_n(\mathcal{A})$ iff $\mathcal{A} \models \sigma_n$.
\end{lemma}

\begin{proof}

\end{proof}

\begin{theorem}
  ${\bf R}(+, ;)$ is axiomatised with the axioms of join semilattice-ordered semigroups and $\{ \sigma_n \}_{n < \omega}$.
\end{theorem}

@@

>>>>

The logic of such structures is the Lambek calculus \cite{lambek1958mathematics} allowing one to characterise inference in categorial grammars, the equivalent version of context-free grammars \cite{pentus1993lambek}. One may define the Lambek calculus as the following (cut-free) sequent calculus.

@@

  \begin{enumerate}
    \item If $\mathcal{A}$ is representable then $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$.
    \item If $\mathcal{A}$ is infinite and $\exists$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is representable.
    \item If $\mathcal{A}$ is infinite and $\forall$ has a winning strategy in $\mathcal{G}_{\omega}(\mathcal{A})$ then $\mathcal{A}$ is not representable.
  \end{enumerate}
\end{lemma}

@@

  \begin{enumerate}
    \item Let $h$ be a representation of $\mathcal{A}$ over some domain $D$. $\forall$ picks $(a, b)$ with $a \not\leq b$ for the initial round. Let $\mathcal{N} = \langle E, V, l \rangle$ be a network being played. To win the game $\exists$ maintains a map $\iota : E \to D$ such that $(\iota(x_0), \iota(x_1)) \in h(b) \setminus h(a)$ and for each $x, y \in E$ and
    $a \in l_{\mathcal{N}}(x, y)$ one has $(\iota(x), \iota(y)) \in h(a)$. The argument repeats the proof of the first item of \cite[Proposition 5]{hirsch2005class}.

    \item
    Firstly, let us suppose that $|\mathcal{A}| = \omega$.

    Let $a, b \in \mathcal{A}$ such that $a \not\leq b$. We consider a play of the game where $\forall$ pick the pair $(a, b)$ in the initial round and plays $(\mathcal{N}, x, y, z, c, d)$ in the further rounds for each $x, y, z \in \operatorname{Nodes}(\mathcal{N})$ and $c,d \in \mathcal{A}$. Here, $c \in l_N(x, y)$ and $d \in l_N(y, z)$.

@@

\item Suppose $(x, y) \in rep(a) ; rep(b)$. We need $(x, y) \in rep(a;b)$. That is, there exists $z \in \operatorname{Nodes}(\mathcal{N})$ with $(x, z) \in rep(a)$ and $(z, y) \in rep(b)$. So, there are $c, d$ such that $c \leq a$ with $c \in l_{\mathcal{N}}(x, z)$ and $d \leq b$ with $d \in l_{\mathcal{N}}(z, y)$.
We also know that $l_{\mathcal{N}}(x, z) ; l_{\mathcal{N}}(z, y) \subseteq l_{\mathcal{N}}(x, y)$ because all composition moves have been played.
So $c ; d \in l_{\mathcal{N}}(x, y)$. That makes $(x, y) \in rep(a;b)$ since $c ; d \leq a ; b$.

\item Injectivity: TODO

\end{itemize}
Now let us assume that $|\mathcal{A}| \geq \aleph_1$. By elementary chain theorem (see \cite[Theorem 2.5.2]{hodges1997shorter})

\item
  \end{enumerate}
\end{proof}

\begin{proposition}
  Let $\mathcal{A}$ be a join semilattice-ordered semigroup, then there exists a game such that $\mathcal{A}$ is representable iff $\exists$ has a winning strategy in the all finite lenght games.
\end{proposition}

  \begin{enumerate}
    \item $(\Rightarrow)$
    \item $(\Leftarrow)$
  \end{enumerate}
\end{proof}

Now we are seeking to obtain the explicit axiomatisation of ${\bf R}(;, +)$. The idea is to axiomatise this class with the additional axioms claiming that $\exists$ has a winning strategy in certain games. See \cite[Chapter 9]{hirsch2002relation} for the discussion in detail to have a more general methodology. Our purpose is to provide the recursively enumerable axiomatisation of ${\bf R}(;, +)$. Thus, according to famous Craig's trick \cite{craig1953axiomatizability}, ${\bf R}(;, +)$ will have the decidable axiomatisation.

\begin{definition}

\end{definition}

@@

\begin{lemma}\label{ax}
  For each $n < \omega$ there exists a first-order sentence $\sigma_n$ such that $\exists$ has a winning strategy in $\mathcal{G}_n(\mathcal{A})$ iff $\mathcal{A} \models \sigma_n$.
\end{lemma}

\begin{proof}
  As usual, for each $n < \omega$ a formula $\sigma_n$ claims that $\exists$ has a winning strategy in the game of lenght $n$.

  Let $\mathcal{N}$ be a term network.

  ...

  We define the sequence of formulas $\{ \sigma_n \}_{n < \omega}$ recursively:
  \begin{enumerate}
    \item $\sigma_0(\mathcal{N}, v) = \bigwedge \limits_{a \in l_{\mathcal{N}}(x, y)} \neg (a \leq v)$

    \item $\sigma_{n + 1} = \sigma_{n} \land $
  \end{enumerate}
\end{proof}

\begin{theorem}
  ${\bf R}(+, ;)$ is axiomatised with the axioms of join semilattice-ordered semigroups and $\{ \sigma_n \}_{n < \omega}$ defined in the proof of Lemma~\ref{ax}.
\end{theorem}

\subsection{The finite representation property for ${\bf R}(+, ;)$}

@@

The following corollary follows from finite representability and finite axiomatisability.

\begin{corollary}
  The class ${\bf R}_{fin}(;, +) = \{ \mathcal{A} \: | \: |\mathcal{A}| < \omega \}$ is recursive. That is, the problem of membership for this class is decidable.
\end{corollary}
